---
title: "ggplot Detailed Features"
subtitle: "for OECD Charte Graphique"
author: ""
date: "2017/06/29"
output:
  xaringan::moon_reader:
#    self_contained: true
    lib_dir: '10-ggplot-detailed-features/libs'
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, center, middle

# theming

http://ggplot2.tidyverse.org/reference/theme.html


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
## Generation of requested POCs shared at http://dki.rdata.work/visualisation/
## Training session to explain concepts employed in POCs (June 2017)
```

```{r xaringan_startup,echo=FALSE,message=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

library(grid)
library(gtable)
## install.packages("cowplot")

library(ggradar)
suppressPackageStartupMessages(library(dplyr))
library(scales)

library(ggiraph)
library(ggiraphExtra)

library(knitr)
opts_chunk$set(digits = 3, tidy = FALSE,
               fig.path = '10-ggplot-detailed-features/fig-html/',
               fig.height = 3.5, fig.width = 10, dev = "svg")
```


---

## Grobs: Legend Width

```{r legend-width-code, tidy=FALSE, echo=TRUE, eval=FALSE}
gp <- ggplot2::ggplotGrob(p)
leg <- gtable::gtable_filter(gp, "guide-box")
panel <- gtable::gtable_filter(gp, "panel")
leg[[1]][[1]]$widths[[elem]] =
  panel$grobs[[1]]$children[[1]]$children[[1]]$width -
  sum(leg[[1]][[1]]$widths) + leg[[1]][[1]]$widths[[elem]]
gp.new = gtable::gtable_add_grob(gp, leg, t = 4, l = 4)
```

```{r legend-width-eval, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
source("../R/ggplot_detailed_features_absLegend.R")
source("../R/ggplot_detailed_features_preparePlot.R")
gp1 <- absLegend(p1)
grid.newpage()
grid.draw(gp1)
```

---

## Legend Guides 

 It is possible to specify the `legend.key` height and width separately for each guide (colour, linetype etc.), see [guide_legend.html](http://ggplot2.tidyverse.org/reference/guide_legend.html)

```{r key-size, tidy=FALSE, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
p1 + guides(fill = guide_legend(keywidth = unit(10, "mm"),
                                keyheight = unit(10, "mm"), title = "",
                                override.aes = list(color = "black")),
            linetype = guide_legend(keywidth = unit(20, "mm"), title = ""))
```

---

## Y-Axis Position

The y-axis title position can be modified in the theme using the parameters of `element_text`, see [theme elements](http://ggplot2.tidyverse.org/reference/element.html)

```{r axis-title-position, tidy=FALSE, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
p1 + scale_y_continuous(name = "This is the y-axis title") +
  theme(axis.title.y = element_text(vjust = 1.08, angle = 0,
                                    margin = margin(r = -75, unit = "pt")))
```

---

## Secondary Y-Axis

The `sec.axis` parameter of `scale_y_continuous` can define a transformed
version of the primary y-axis,
see [sec_axis()](http://ggplot2.tidyverse.org/reference/sec_axis.html). Prior to
plotting, series with different units need to be scaled to the same unit manually.

```{r sec-axis, echo=TRUE, cache=TRUE}
ggplot(mtcars, aes(cyl, mpg)) + geom_point()
```

---

We may want to display the levels of `mpg` and `hp` on the same plot

```{r mtcars-datatable, eval=require('DT'), echo=FALSE, cache=TRUE}
DT::datatable(
  head(mtcars, 10),
  fillContainer = FALSE,
  options = list(pageLength = 5)
)
```

```{r mtcars-summary, echo=TRUE}
mtcars %>% select(mpg, hp) %>% mutate(hp_scaled = hp / 10) %>% summary()
```

---

Plot on separate axes, manual axis title and legend labels

```{r rescale-mtcars, cache=TRUE}
scalefactor <- 10^1
mtcars %>%
  mutate(hp_scaled = hp / scalefactor) %>%
  select(cyl, mpg, hp_scaled) %>%
  gather(key = variable, value = value, -cyl) %>%
  ggplot(aes(x=cyl, y = value, color = variable)) +
  geom_point() +
  scale_y_continuous(name = "mpg",
                     sec.axis = sec_axis(name="hp", ~.*scalefactor)) +
  scale_color_discrete(labels = c("hp", "mpg"))
```

---
class: inverse, center, middle

# aesthetics

---

## data labels

The `geom_text` layer accepts parameters for `element_text`, for details see
[geom_text](http://ggplot2.tidyverse.org/reference/geom_text.html).
The fonts must be installed on your computer.

```{r pie-table, echo=FALSE, message=FALSE, cache=TRUE}
df <- data.frame(categ = c(1, 2, 3, 4),
                 pct = c(27.82, 28.70, 22.44, 21.04))
df_lab <- df %>% mutate(labelpos = 100 - (cumsum(pct) - pct / 2))

p <- ggplot(df_lab, aes(x = factor(0), y = pct, fill = factor(categ))) +
  geom_bar(width = 1, stat = "identity", color = "black", size = 0.25) +
  theme(legend.position = "none") +
  scale_x_discrete(NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, expand = c(0, 0)) +
  scale_fill_manual(
    values = c("#929292", # dark grey
               "#a6b8e2", # light blue
               "#cbcbcb", # light grey
               "#4e80bc"  # dark blue
               ))
```

```{r pie-label, tidy=FALSE, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
## labelpos = 100 - (cumsum(pct) - pct / 2)
p + geom_text(aes(y = labelpos,
                  label = paste0(as.character(pct), "%")),
              family = "serif", size = 6, color = "red") +
  coord_polar(theta = "y")
```

---

## High Low Plot

Analogue to colors, linetypes and shapes can be mapped to variables,
see [Shapes and line types](http://www.cookbook-r.com/Graphs/Shapes_and_line_types/)


```{r high-low-data, echo=FALSE, message=FALSE, cache=TRUE}
labels_hl <- data.frame(variable = c("h", "m", "l"),
                        label = c("High Price", "Middle Price", "Low Price"),
                        stringsAsFactors = FALSE)
n <- 14
sd <- 0.6
df_hl <-
  data.frame(
    date = as.Date(paste0("2011-01-", as.character(seq(1:n)))),
    m = rnorm(n, mean = 0.8, sd = 0.2)
  ) %>%
  mutate(h = m + abs(rnorm(n, sd = sd)),
         l = m - abs(rnorm(n, sd = sd))) %>%
  gather(key = variable, value = value, -date) %>%
  left_join(labels_hl, by = "variable")
df_hl$label <- factor(df_hl$label, levels = labels_hl$label)

themeHighLow <- function(p, col=c("black", "#4f81bd", "black")) {
  p1 <-
    p +
    scale_color_manual(values = col) +
    scale_fill_manual(values = col) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      axis.line.x = element_line(color = 'black')
    ) +
    xlab(NULL) + ylab(NULL)
  return(p1)
}

```

```{r high-low-plot-base, echo=TRUE, message=FALSE, cache=TRUE}
p_hl <- ggplot(data = df_hl, aes(x = date, y = value,
                                 color = label, fill = label)) +
  geom_point(aes(shape = label), size = 2) +
  scale_shape_manual(values = c(17, 23, 15))
themeHighLow(p = p_hl, col = c("black", "#4f81bd", "black"))
```

---

## High Low Plot with Lines

`geom_segment` allows drawing arbitrary lines, see [Line segments and curves](http://ggplot2.tidyverse.org/reference/geom_segment.html)

```{r high-low-plot-line, echo=TRUE, message=FALSE, cache=TRUE}
df_hl2 <- df_hl %>% group_by(date) %>% mutate(min=min(value), max=(max(value))) %>% ungroup()
p_hl2 <- ggplot(data = df_hl2, aes( x = date, y = value, color = label, fill = label )) +
  geom_point( aes( shape = label ), size = 2) +
  geom_segment( aes( xend = date, y = min, yend = max ), size = 0.05) +
  scale_shape_manual(values = c(17, 23, 15)); themeHighLow(p = p_hl2)
```


---
class: inverse, center, middle

# extensions

---

## Radar Plot

Data extracted from http://stats.oecd.org/Index.aspx?DataSetCode=BLI

```
mutate_at(vars(-LOCATION), funs(scales::rescale))
```

```{r radar-data, message=FALSE, echo=FALSE, cache=TRUE}
bli_radar <-
  "../data/BLI_2016.csv" %>%
  read.csv() %>%
  filter(INEQUALITY == "WMN") %>%
  select(LOCATION, Indicator, Value) %>%
  spread(key = Indicator, value = Value) %>%
  mutate_at(vars(-LOCATION), funs(scales::rescale)) %>%
  filter(LOCATION %in% c("FRA", "ITA")) %>%
  rename(group = LOCATION)

```

```{r radar-table, echo=FALSE, results='as.is'}
## names(bli_radar)
## knitr::kable(bli_radar[, c(1,5:6)])
str(bli_radar)
```

---

## Radar Plot

The existing ggplot2 extension [ggradar](http://www.ggplot2-exts.org/ggradar.html) can be modified. Alternatively, the [plotrix package](https://cran.r-project.org/web/packages/plotrix/index.html) can be tested. More examples can be found [here](http://boot.rdata.work/contrib/radar-example/)

```{r radar-chart-1, message=FALSE, warning=FALSE, echo=TRUE, cache=TRUE}
p_radar <-
  ggradar(bli_radar,
          font.radar = "sans",
          grid.label.size = 4,
          axis.label.size = 4,
          group.point.size = 1,
          group.line.width = 1,
          plot.extent.x.sf = 1.8) +
  scale_color_manual(values = c("#8cc841", "#da2128")) +
theme(
  legend.position = c(0.05, 0.95),
  legend.justification = "left",
  legend.direction = "horizontal",
  legend.text = element_text(size = 8),
  legend.background = element_rect(fill = NULL)
)
```

---

## Radar Plot

```{r radar-chart-2, fig.width=11.5, fig.height=5.5, message=FALSE, warning=FALSE, echo=FALSE, cache=FALSE}
p_radar
## plotly::ggplotly(p_radar)
```

